---
title: "SOCI 30005_PS3_Hinojosa"
author: "Cintia Hinojosa"
date: "5/19/2019"
output:
 pdf_document:
 keep_tex: true
---

## Setup
**Git-Overleaf publish commands**
git pull origin master
git add .
git commit -m “message” 
git push -u origin master
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	error = TRUE,
	warning = FALSE
)
```

```{r libraries, message=FALSE, warning=FALSE, include=FALSE} 

# data files
library(readr)
library(foreign)
library(haven)
library(knitr)

# data wrangling
library(tidyverse)
library(plyr)
library(dplyr)
library(tidyimpute)
library(magrittr) # %>%
library(qwraps2) # wrapper
library(reshape2) 
library(forcats) # factor variables
library(doBy) # groupwise 
library(fastDummies)  # dummy_cols(.data, select_columns = NULL)

# summaries
library(stargazer) 
library(summarytools)
library(jtools) # summ(): summarize/visualize regressions
library(naniar) # summarize & missing data; w/tidy
library(huxtable) # format text tables

# analyses
library(lmtest)
library(psych)
library(rdrobust) # regression discontinuity
library(gmodels) # glm parameters, estimates, contrasts, tests 

# plots
library(ggplot2)
library(HistogramTools)

```

```{r load}
setwd("C:/Users/cinti/Box Sync/Booth 2017-2018/Spring 2019/SOCI 30005/")

csiw <- read_sav("csiw_final.sav")

summary(csiw)
```

### Key Variables
- CSIW (treatmt)
    - 1=CSIW
    - 0=control
- Achievement Level (group)
    - 1=High
    - 2=Average
    - 3=Low
    - 4=Learning Disability 
- Holistic pretest (cch1)
    - pre-test on writing achievement
- Holistic posttest (cch2)
    - post-test on writing achievement
- Grade (grade)
    - 1=Grade 4
    - 2=Grade 5
- School

## Intro

**Our aim in this assignment is the same as it was in Assignment 2: To study the impact of cognitive strategies on writing on writing (CSIW). But now we are going to confront and solve two key problems that were ignored in Assignment 2: a) the nested character of the data; b) missing data.**

**You should have already re-coded Achievement Level into 4 dummy variables and grade into 1 dummy variable.**
```{r recode}

# achievement level into 4 dummy variables

# grade into 1 dummy variable

```


## A. Nested Data 
**Run a cross tab in which the rows are the schools (“school”) and rows are CSIW. Based on this, tell us at what level the treatment (“CSIW”) varies.** 

## B. Build the MDM file 
**Sort the cases by “school.” The level-1 and level-2 files will be the same. Choose the relevant child level variables at level-1 and the relevant school variable at level 2. B. Analysis of covariance (ANCOVA) model. Indicate that there are missing data, and tell the program to delete missing level-1 cases at run time.**


## C. Analysis of Covariance
**1. Write down the level-1 model with relevant covariates (do not include quadratic terms or interactions at this time). 2. Write down the level-2 model. 3. Estimate the model and tell us the estimated treatment effect and its standard error (model based and robust).**