---
title: "PS2"
author: "Cintia Hinojosa"
site: bookdown::bookdown_site
documentclass: book
output:
  bookdown::gitbook: default
  bookdown::pdf_book: default
---

# Intro

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r libraries, message=FALSE, warning=FALSE, include=FALSE} 
library(haven)
library(tidyverse)
library(dplyr)
library(readr)
library(codebook)
library(psych)
library(gmodels)
library(doBy)
library(rdrobust)
library(foreign)
library(ggplot2)
library(plyr)
library(stargazer)
library(lmtest)
library(gmodels)
library(readr)
```

```{r load}
csiw <- read.spss("csiw07.sav", to.data.frame = TRUE)
```

## CSIW Dataset
- CSIW (treatmt)
    - 1=CSIW
    - 0=control
- Achievement Level ()
    - 1=High
    - 2=Average
    - 3=Low
    - 4=Learning Disability 
- Holistic pretest 
    - pre-test on writing achievement
- Holistic posttest ()
    - post-test on writing achievement
- Grade (grade)
    - 1=Grade 4
    - 2=Grade 5
```{r csiw dataset}

names(csiw)
summary(csiw)
codebook(csiw)
```

# Naive Model

# ANCOVA Model

# Assessing Linearity

# Quadratic Model

# Searching for Confounders

# Heterogeneity

# Checking Assumptions

# Conclusion

