---
title: "Intro to R Week 2 Lab 1"
params:
  dotabs: no
output:
  pdf_document: default
  html_document:
    df_print: paged
utput:
  html_document:
    code_folding: hide
    css: styles.css
    df_print: kable
    number_sections: yes
    self_contained: yes
    theme: cerulean
  pdf_document: default
---

### R Studio Interface
The R interface has 4 panels which include:  
- scripts (markdown documents, notebooks)  
- console (command line)  
- enviornment (lists of created/availabe objects)  
- files/plots/viewer  
    
#### Packages  
You use packages to run your code. For any package, you install just once, then use `library` to load the package when needed.  
For example:  

`install.packages("packagename")`  You only need to do this once.
You can also install packages by using the interface:    
- Toos > Install Packages   
To envoke a package (which you need to do everytime you start working in R)  
`library(packagename)` when you start up RStudio.
    
#### Working Directory  
- where you keep your files, data, etc  
- always know where you are!   
- make a seperate workign directory for each project  
  
#### Working Directory Basics
  * get working directory
- set working directory  
- or use the File panel in R studio  
  
`getwd()`  
`setwd()`  

# Computational Thinking  

#### Algorithms and Data Structures  
- formal steps to solve a problem  
  
#### Data Structures  
- how to store information in a computer  
- everything is an object  
  
#### Types of Data Structures  
- data types  
   `character`: for text e.g., "statistics"  
   `numeric`: e.g., 219.2  
   `integer`: e.g., 44  
   `logical`: TRUE , FALSE  
   `factor`: for categories   
   
#### What is the Data Types  
- what is an object?       
      `is.character`  
      `is.numeric`  
      `is.logical`  
      
```{r}
is.character("2")
is.character(2)

is.numeric(2)
is.numeric("2")

is.logical(1)
is.logical(TRUE)
```
#### Converting the Data type  
- convert one type into another  
```{r}
as.numeric("2")
as.character(2)
```
  
#### Assigning Variables   
- everything in R is an object   
- create objects by assigning values to them   
- assign the value of 5 to x  
- lets see what it is assigns the value to  


```{r}
x<-5
x
mode(x)
is.character(x)
is.integer(x)
is.numeric(x)

```
#### Mathematical Operations  
  + - * /      
  %% modulus or remainder  
  %/% integer division  
  ^  power  
```{r}
5+5
10-5
5*5
25/5

7%%2
11 %/% 2
3^2

a<-2
b<-3
a+b
a*b
b^a
c<-a+b
c
d<-a*b
d
e<-c+d
e
is.numeric(e)
d>c
f<-c==d
f
is.logical(f)
g<-as.numeric(f)
g
```
#### Operators   
  ==   
  all.equal()  
  >   
  <  
  >=  
  <=  

```{r}
5>4
5<4
5<=5
5>=5
```

### Program Flow  
1) get data and assign to data structures  
2) move through steps in the algorithm  
    logical branches: if statments  
    repeated operations: for loops  
3) combine operations into functions  

#### Common Mistakes in R  
common mistakes in R  
- using wrong case  
- forgetting ""  
- forgetting ()  
- using function from a package that is not loaded  
- typos

# Basic Data Structures  
#### Vector  
- collection of values of the same type  
- concatentate   
- c(elements)  
```{r}
x <- c(3,7,4)
x
length (x)

z <- c("a","b","c")
z
length(z)
```
- sequences   
- seq(n1,n2,increment)  
- n1:n2  
```{r}
y <- seq(1,10)
y
y2 <- seq(1,10,2)
y2
```
#### Extracting Vector Elements  
- specific element   
  `[index]`  
```{r}
x<- seq(1,10)
x
x[4]
x[1:5]
```
#### Matrices   
- regular, two dimension tables  
- convert a vector to matrix (x, nrow=n, ncol=m)  
- dimension of matrics  
    `dim`  
```{r}
v1<-1:16
v1

v2 <- matrix(v1, nrow=4,ncol=4)
v2

v3 <- matrix(v1, nrow=4,ncol=4,byrow=TRUE)
v3
dim(v3)
```
#### Extracting Matrix Elements   
- finding a point in the matrix [rowselection, columnselection]  
```{r}
v3
v3[2,2]
v3[1:2,3:4]
v3[,2]
v3[,3:4]
v3[c(1,3),]
```
#### Lists  
- collection of elements of different types  

```{r}
z <- list (name="point", x=3.5, y=2)
z
```

#### Data Frames  
- columns= variables  
- rows= observations  
```{r}
v1 <- 1:16
m1 <- matrix(v1, nrow=4,ncol=4)
m1

t1 <-data.frame(m1)
t1

t2 <-as.data.frame(m1)
t2


```
#### what is the difference between these two??  
```{r}
names(t1)
row.names(t1)
names(t2)
row.names(t2)
```
# Lab 2 Week 3  
# Reading a Data Frame from a Text File    
  `read.table`  
  `read.csv`  
### things to watch out for  
- make sure file is in working director  
- header (sets the variable names)  
- set stringsAsFactors=FALSE  
  
### Reading a data file  
  data <- read.csv("NSYL.csv")  
    
### some useful commands  
  view the first 6 rows  
    head (data)  
  structure, gives types of columns  
    str  
  dimension, number of rows and columns  
    dim  
  statistical summary of variables: min, 1st quartile, median, mean, 3rd quartils, max  
    summary  
    
### Extracting variables  
  sometimes you will want to extract a vector to look at just a subset of data  
  
### useful commands  

#### extracting variables columns   
  data_frame$variable_name  
  [["variable,name"]]  
  [,column_number]  
  
#### extracting a data frame  
  [column_number]  
  
#Writing a Data Frame to a file  
  write.csf(df,filename)  
    set row.names=FALSE  
  will be written to workign directory  
 
  